# Agricultural and Forestal Districts

Local governments are permitted to enact an ordinance providing for the creation of agricultural and forestal districts. Such districts are intended, as the *Code* states, "... to conserve and to encourage the development and improvement of the commonwealth's agricultural and forestal lands for the production of food and other agricultural and forestal products." According to the *Code*, the districts also "...conserve and protect agricultural and forestal lands as valued natural and ecological resources which provide essential open spaces for clean air sheds, watershed protection, wildlife habitat, as well as for aesthetic purposes." The authority for such districts is provided by the *Code of Virginia*, §§ 15.2-4300 through 15.2-4314 (Agricultural and Forestal Districts Act) and §§ 15.2-4400 through 15.2-4407 (Local Agricultural and Forestal Districts Act). 

In accordance with the Agricultural and Forestal Districts Act, each district must have a core of no less than 200 acres in one parcel or in contiguous parcels; however, districts of local significance created under the act may be as small as 20 acres.[^05-1] Further, the local governing body must review each district within four to ten years after its creation and every four to ten years thereafter. For additional information relating to the creation of the districts, see § 15.2-4305. 

Land devoted to agricultural and forestal production within an agricultural and forestal district qualifies for special assessment for land use whether or not a local land use plan or special assessments ordinance has been adopted, provided that the land meets the criteria set forth in §58.1-3230 et seq. of the *Code* (see also § 15.2-4312).[^05-2]

Three cities and 28 counties reported having a total of 372 agricultural and forestal districts. In addition, two towns, Blacksburg and Louisa, reported a total of two districts. In terms of acreage, Cities reported a total of 2,530 acres and the two towns reported a total of 1,422 acres - 1,360 acres and 62 acres, respectively. These numbers were negligible compared to the 736,140 acres reported by counties. Of the counties, those reporting the greatest number of acres within agricultural and forestal districts were Fauquier (78,755 acres), Accomack (74,093 acres), Albemarle (72,665 acres), Culpeper (46,487 acres), and Isle of Wight (41,317 acres). 

The following text table shows by year when the existing city and county districts came into existence. Four new districts were reported in 2019. 

```{r}
# table name: Agricultural and Forestal Districts by Year of Creation for Cities and Counties, 1978 and 2019
```

**Table 5.1** presents information for all cities, counties, and towns which reported agricultural and forestal districts. The table includes the district creation date, acreage, and the review period for each district. Three cities, 28 counties and two towns reported having an agricultural and forestal district ordinance in effect for the 2019 tax year. 

Section 4 of this publication provides details on the related program of land use value assessments and cites literature that questions the effectiveness of special assessments in slowing the conversion of participating land to other uses. 

```{r}
# Table 5.1 Agricultural and Forestal Districts, 2019
```

```{r load code to make tables 20, message=FALSE, echo=FALSE}
source("make_table.R")
```

```{r table05-1, echo=FALSE, eval=FALSE}
#table_05_1_vars   <- c("ExternalDataReference", "")

#table_05_1_labels <- c("Locality","Name of District", "Date Created", "Review Period (Years)", "Acreage") 

#table_05_1 <- make_table(table_05_1_vars)

#knitr::kable(table_05_1, 
             #caption = "Agricultural and Forestal Districts Table, 2019", 
             #col.names = table_05_1_labels, 
             #align = "lcccc",
             #format = "html")
```


[^05-1]: Under provisions of the Local Agricultural and Forestal Districts Act, only counties satisfying the following conditions are "participating localities" empowered to establish districts with this reduced acreage requirement: (1) a county with an urban county executive form of government, (2) any adjacent county having the county executive form of government, and (3) counties with population sizes ranging from 63,400 to 73,900 or from 85,000 to 90,000 [no census cited]. See §§ 15.2-4402 through 4405. 


[^05-2]: For additional rules concerning agricultural and forestal districts, see § 15.2-4312. 
