# Taxes on Natural Resources

```{r prep data for section 17, message=FALSE, echo=FALSE}
source("get_and_prep_data.R")

section_vars <- c("nat_tax_coal",
                  "nat_tax_oil",
                  "nat_tax_add_gas",
                  "nat_tax_road_impr",
                  "nat_tax_min")

tax_rates %>% 
  select(all_of(reference_vars), all_of(section_vars))  %>%
  mutate(across(.cols = all_of(section_vars), ~ ifelse(.==0, NA, .)))   %>%   # Convert zero values to NA's
  mutate(across(.cols = all_of(section_vars), ~ !is.na(.), .names = "has_{.col}")) %>%
  mutate(has_nat_tax_min = (nat_tax_min == "Yes")) %>%
  filter(if_any(starts_with("has_"))) -> section_tbl



```

Taxes on natural resources  are  rarely  used  by  localities  because many are not endowed with such resources. As a consequence, natural resources taxes accounted for less than 0.1 percent of total city tax revenue in fiscal year 2018, 0.2  percent of total county tax revenue, and less than 0.1 percent of total tax revenue of large towns, according to information from the Auditor of Public Accounts. These are averages; the vast majority of localities receive no revenue from this source. All the exceptions are located in Southwest Virginia. For information on individual localities, see Appendix C.

Localities  are  permitted  to  impose  several  taxes  on  natural resources. **Table 17.1** provides tax rates for the cities and counties having such natural resource-related taxes in effect during the 2019 tax year.

## TAXATION OF MINERAL LANDS

Under § 58.1-3286 of the *Code of Virginia*, localities are required to “...specially and separately assess at the fair market value all mineral lands and the improvements thereon...” and enter those assessments separately from assessments of other lands and improvements. Mineral lands are taxed at the same rate as other real estate in the locality. Localities may request technical assistance from the Virginia Department of Taxation in assessing mineral lands and minerals, provided money is available to the department to defray the cost of the assistance (§ 58.1-3287). Instead of employing the real property tax for mineral lands, localities are permitted to substitute a severance tax on mineral sales, not to exceed 1 percent.

In 2009, this section was amended to allow Buchanan County to reassess mineral  lands  on  an  annual  basis  for  purposes of determining the real property tax on such land. Other real estate is still subject to assessment every six years. Currently, 2 cities and 23 counties report assessing taxes on minerals. Among the several that commented on their mineral tax, most stated they used the land assessment method. The city of Norton, however, stated that its tax was based on a loading tax of $0.05/ton.

## SEVERANCE TAX

Under § 58.1-3712, any city or county may levy a license tax on businesses engaged in severing coal and gases from the earth. The maximum rate permitted is 1 percent of the gross receipts from sales. A 2012 bill reduced the rates of the local coal severance tax for small mines from 1 percent to 0.75 percent of the gross receipts from the sale of  coal. “Small mine” is defined here as a mine that sells less than 10,000 tons of coal per month. 

Localities choosing to use § 58.1-3712 may not exercise the option to levy a 1 percent severance tax under § 58.1-3286.  Under  §  58.1-3712.1,  the  maximum  rate  permitted  for severing oil is one-half of 1 percent from the sale of the extracted oil. Notwithstanding the rate limits established in §  58.1-3712,  cities  or  counties  may  impose  an  additional  license tax of 1 percent of the gross receipts from the sale of gas severed as authorized by § 58.1-3713.4. The funds from this additional levy are paid into the general fund of the localities except for members of the Virginia Coalfield Economic Development Fund, where one-half of the revenues must be paid to the fund. The members of the fund are the counties of Buchanan, Dickenson, Lee, Russell, Scott, Tazewell, and Wise and the city of Norton. 

## COAL AND GAS ROAD IMPROVEMENT TAX

Notwithstanding  the  rate  limits  described  in  the  previous  paragraph, localities are permitted by § 58.1-3713 to levy up to an additional 1 percent license tax on the gross receipts of coal and gas extracted from the ground. As with the severance tax on coal, the coal road improvement tax has been modified to reduce the tax from 1 percent to 0.75 percent for small mines. This tax was originally scheduled to end in 2007, but the General Assembly extended the sunset clause a number of times, most recently to December 31, 2017.

The  amount  collected  under  this  tax  must  be  paid  into  a  special  fund  to  be  called  the  Coal  and  Gas  Road  Improvement Fund of the particular county or city where the  tax  is  collected.  In  addition,  “the  county  may  also,  in  its discretion, elect to improve city or town roads with its funds  if  consent  of  the  city  or  town  council  is  obtained.”  One-half  of  the  revenue  paid  to  this  fund  may  be  used  for  the  purpose  of  funding  the  construction  of  new  water  systems  and  lines  in  areas  of  insufficient  natural  supply  of water. Those same funds may also be used to improve existing water and sewer systems. Localities are required to develop and ratify an annual funding plan for such projects. Under § 58.1-3713.1, 20 percent of the funds collected in Wise County are distributed to the six incorporated towns within  the  county’s  boundaries  (Appalachia,  Big  Stone Gap,  Coeburn,  Pound,  Saint  Paul,  and  Wise)  and  the  city  of  Norton.  The  distribution  is  determined  as  follows:  (a)  25  percent  is  divided  among  the  incorporated  towns  and  the city of Norton based on the number of registered motor vehicles  in  each  town  and  the  city  of  Norton,  and  (b)  75  percent is divided equally among the towns and the city of Norton.  The  Coal  and  Gas  Road  Improvement  Advisory  Committee in the city of Norton and county must develop a plan  before  July  1  of  each  year  for  road  improvements for the following fiscal year. For localities in the Virginia Coalfield  Economic Development  Authority  (Lee,  Wise,  Scott, Buchanan, Russell, Tazewell, and Dickenson counties and the city of Norton), the receipts from this tax are distributed  as  follows:  three-fourths  to  the  Coal  and  Gas  Road  Improvement  Fund  and  one-fourth  to  the  Virginia  Coalfield Economic Development Fund. The purpose of this fund is to enhance the economic base for the seven counties and one city in the authority. 


```{r table17-1, echo=FALSE}
# table_17_1 <- section_tbl %>% select(locality_name, all_of(section_vars))

table_17_1 <- section_tbl %>% select(locality_name, locality_group, all_of(section_vars))

table_17_1_labels <- c("Locality",
                       "Coal & Gas Severance Tax\\\n(§ 58.1-3712)",
                       "Oil Severance Tax\\\n(§ 58.1-3712.1)",
                       "Additional Gas Severance Tax\\\n(§ 58.1-3713.4)", 
                       "Coal & Gas Road Improvement Tax\\\n(§ 58.1-3713)", 
                       "Tax on Mineral Land\\\n(§ 58.1-3286)")

# kbl(tbl %>% select(-locality_group), 
#              caption = "Natural Resource Taxes, 2019", 
#              col.names = table_17_1_labels, 
#              align = "lccccc",
#              format = "html") %>%
#   pack_rows(index = count_localities_by_type(tbl))

make_long_table(table_17_1, 
             caption = "Natural Resource Taxes, 2019", 
             col.names = table_17_1_labels, 
             align = "lccccc")
```
