# Refuse and Recycling Collection Fees

```{r prep data for section 20, message=FALSE, echo=FALSE}
#source("get_and_prep_data.R")

section_vars <- c("RCC_tipping_fees",
                  "RCC_rec_serv_type",
                  "RCC_rec_charge_fee")

#tax_rates %>% 
  #select(all_of(reference_vars), all_of(section_vars))  %>%
  #mutate(across(.cols = all_of(section_vars), ~ ifelse(.==0, NA, .)))   %>%   # Convert zero values to NA's
  #mutate(across(.cols = all_of(section_vars), ~ !is.na(.), .names = "has_{.col}")) %>%
  #mutate(has_RCC_tipping_fees = case_when(!(has_RCC_tipping_fees) ~ FALSE,
                                          #RCC_tipping_fees %in% c("no", "No", "N/A", "Landfill closed") ~ FALSE,
                                          #TRUE ~ TRUE)) %>%
  #mutate(has_RCC_rec_serv_type = (RCC_rec_serv_type != "Not Applicable")) %>%
  #filter(if_any(starts_with("has_"))) -> section_tbl
```

Many Virginia localities collect, or authorize to have collected,  refuse  and  recycled  materials.  In  its  survey,  the  Cooper Center inquired into the methods and fees for the collection  of  refuse  and  recycled  materials.  The  answers  are provided in four tables covering regular refuse pick up, tipping fees, recycling, and pickup of miscellaneous refuse items.

## REFUSE COLLECTION

**Table 20.1** shows information reported on refuse collection by all 38 cities, and by 24 counties and 101 towns. The table contains information on frequency of collection, collection fees  and  private  contracting.  There  are  three  methods  of  operation.  Some  Virginia  localities  levy  a  specific  refuse collection service fee for the costs of collection. Others pay for collection costs with general tax revenues. Finally, some localities provide no service; instead, they leave refuse collection to private contractors. 

A  majority  of  cities  and  counties  provide  basic  residential  services  on  a  weekly  basis.  Only  the  counties  of  Arlington, Chesterfield, and Halifax offer regular collections more frequently.

Regarding  fees,  32  cities,  13  counties,  and  62  towns  reported  imposing  a  residential  refuse  collection  service  fee. Eleven cities, 7 counties, and 42 towns contracted with private firms  for  refuse  collection.  The  text  table  below  shows this breakdown.

```{r}
#text table "Residential Refuse Collection, 2019" goes here
```

**Table 20.2** shows tipping fees charged by various localities  to  dump  trash  at  landfills  and  waste  transfer  stations.  Localities  reporting  imposing  such  fees  included  9  cities,  34 counties, and 7 towns.

## RECYCLING PROGRAMS

**Table 20.3** provides data on localities that have instituted recycling  programs.  As  with  refuse  collection,  these  programs may be financed in a variety of ways. Many localities pick  up  recyclables  and  then  finance  the  collection  with  a  service  charge.  Other  localities  contract  with  a  private  firm. **Table 20.3** shows which localities offer collection of recyclables and which contract for collection with a private firm. It also shows the monthly fees associated with collecting recyclables.

```{r Residential Recycling Programs, include=FALSE}
#section_tbl %>%
  #group_by(locality_group) %>%
  #summarize(Direct     = sum(RCC_rec_serv_type == "Directly"),
            #Contracted = sum(RCC_rec_serv_type == "Contracted")) %>%
  #ungroup() %>%
  #pivot_longer(!locality_group, names_to = "Service", values_to = "count") %>%
  #pivot_wider(names_from = locality_group, values_from = count) -> res_recycling_summary_tbl
```

Of  the  total  survey  respondents,  38  cities,  83  counties, and 67 towns reported having some form of recycling activity.  Seventeen  cities  provided  recycling  collection  directly,  and  21  contracted  it  out.  Thirty-seven  counties  provided services directly, while 46 contracted them out. Of the towns, 8 had their services provided by their host county, 25 provided direct services, and 34 contracted for services. The table below shows this breakdown.


```{r , echo=FALSE}
#res_recycling_summary_tbl %>%
  #knitr::kable(caption = "Residential Recycling Programs, 2019")
```


For  localities  that  charged  a  service  fee,  the  amount  ranged anywhere from \$1.33 to $16.50 per month.

```{r}
#Table 20.1 "Refuse Collection Fees, 2019" goes here, formatting TBD
```

```{r load code to make tables 20, message=FALSE, echo=FALSE}
source("make_table.R")
```

```{r table20-1, echo=FALSE, eval=FALSE}
table_20_1_vars   <- c("ExternalDataReference", "RCC_res_freq","RCC_ind_freq", "RCC_res_fee", "RCC_ind_fee", "RCC_private_contract")

table_20_1_labels <- c("Locality","Frequency of Collections per Week", "", "Collection Fee (Dollars Per Month Unless Otherwise Stated)", "", "Locality Contracts with Private Firm(s) for Refuse Collection")

table_20_1 <- make_table(table_20_1_vars)

knitr::kable(table_20_1, 
             caption = "Refuse Collection Fees Table, 2019", 
             col.names = table_20_1_labels, 
             align = "lccc",
             format = "html")
```

```{r table20-2, echo=FALSE}
#section_tbl %>% 
  #filter(has_RCC_tipping_fees) %>%
  #select(locality_name, locality_group, RCC_tipping_fees) -> table_20_2

#table_20_2_labels <- c("Locality","Tipping Fee")

#make_long_table(table_20_2, 
             #caption = "Refuse Collection Tipping Fees Table, 2019", 
             #col.names = table_20_2_labels, 
             #align = "lc")
```

```{r table20-3, echo=FALSE}
#section_tbl %>% 
  #filter(has_RCC_rec_serv_type) %>%
  #select(locality_name, locality_group, RCC_rec_serv_type, RCC_rec_charge_fee) -> table_20_3

#table_20_3_labels <- c("Locality","Provided Directly or Contracted","Service Fee")

#make_long_table(table_20_3, 
             #caption = "Recycling Collection Fees Table, 2019", 
             #col.names = table_20_3_labels, 
             #align = "lcc")
```
