--- 
title:    "Virginia Local Tax Rates, 2019"
subtitle: "38th Annual Edition: Information for All Cities and Counties and Selected Incorporated Towns"
author:   "Stephen C. Kulp, Weldon Cooper Center for Public Service, University of Virginia"
date:     "2021"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "Information for All Cities and Counties and Selected Incorporated Towns"
---

# Overview



## Credits

These course materials were generated using the **bookdown** package [@R-bookdown], which was built on top of R Markdown and **knitr** [@xie2015].

<!--chapter:end:index.Rmd-->


# Introduction

Placeholder


## Foreward
## Organization of the Book
## About the Survey
## Some Components of Local Taxes
## Partnership with LexisNexis
## Study Personnel
## Final Comments

<!--chapter:end:va-local-tax-rates-introduction.Rmd-->


# Section 1

Placeholder


## Summary of Legislative Changes in Local Taxation, 2019 $^1$
### General Provisions
#### Local License Tax on Mobile Food Units
#### Local Gas Road Improvement Tax; Extension of Sunset Provision
#### Private Collectors Authorized for Use by Localities to Collect Delinquent Debts
### Real Property Tax
#### Real Property Tax Exemptions for Elderly and Disabled: Computation of Income Limitation
#### Real Property Tax Exemption for Elderly and Disabled: Improvements to a Dwelling
#### Real Property Tax: Partial Exemption from Real Property Taxes for Flood Mitigation Efforts
#### Real Property Tax: Exemption for Certain Surviving Spouses
#### Land Preservation; Special Assessment, Optional Limit on Annual Increase in Assessed Value
#### Virginia Regional Industrial Act: Revenue Sharing; Composite Index
#### Real Estate with Delinquent Taxes or Liens: Appointment of Special Commissioner; Increase Required Value
#### Real Estate with Delinquent Taxes or Liens; Appointment of Special Commissioner in the City of Martinsville
### Personal Property Tax
#### Constitutional Amendment: Personal Property Tax Exemption for Motor Vehicle of a Disabled Veteran
#### Personal Property Tax Exemption for Agricultural Vehicles
#### Intangible Personal Property Tax: Classifi cation of Certain Business Property

<!--chapter:end:va-local-tax-rates-Section-1.Rmd-->


# Section 15

Placeholder


## Motor Vehicle Local License Tax, 2019

<!--chapter:end:va-local-tax-rates-Section-15.Rmd-->


# Section 16

Placeholder


## Meals, Transient Occupancy, Cigarettes, Tobacco, and Admissions Excise Taxes, 2019
### MEALS TAX
### TRANSIENT OCCUPANCY TAX
### CIGARETTE AND TOBACCO TAXES
### ADMISSIONS TAX

<!--chapter:end:va-local-tax-rates-Section-16.Rmd-->


# Section 17

Placeholder


## Taxes on Natural Resources, 2019
### TAXATION OF MINERAL LANDS
### SEVERANCE TAX
### COAL AND GAS ROAD IMPROVEMENT TAX

<!--chapter:end:va-local-tax-rates-Section-17.Rmd-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Section 18

## Legal Document Taxes, 2019
|   In fiscal  year  2018,  the  most  recent  year  available  from  the Auditor of Public Accounts, taxes on legal documents  accounted for 0.5 percent of total tax revenue for cities and 0.8 percent for counties. Towns do not have this tax. These are averages; the relative importance of taxes in individual localities may vary significantly. For information on individual localities, see Appendix C.
|
|   Section 58.1-3800 of the Code of Virginia authorizes the governing body of any city or county to impose a recordation tax in an amount equal to one-third of the state recordation tax. The recordation tax generally applies to real and personal property in connection with deeds of trust, mortgages, and leases, and to contracts involving the sale of rolling stock or equipment (§§ 58.1-807 and 58.1-808). 
|
|   Local governments are not permitted to impose a levy when the state recordation tax imposed is 50 cents or more (§  58.1-3800).  Consequently,  local  governments  cannot  levy a tax on such documents as certain corporate charter amendments (§ 58.1-801), deeds of release (§ 58.1-805), or deeds of partition (§ 58.1-806) as the state tax imposed is already 50 cents per $100. 
|
|   Sections  58.1-809 and 58.1-810 also specifically exempt certain types of deed modifications from being taxed. Deeds of confirmation or correction, deeds to which the only parties are husband and wife, and modifications or supplements to the  original  deeds  are  not  taxed.  Finally,  §  58.1-811  lists  a  number of exemptions to the recordation tax. 
|
|   Currently,  the  state  recordation  tax  on  the  first  \$10 million of value is 25 cents per \$100, so cities and counties can  impose  a  maximum  tax  of  8.3  cents  per  \$100  on  the  first \$10 million, one-third of the 25 cent state rate. Above $10 million there is a declining scale of charges applicable (§ 58.1-3803).
|
|   In  addition  to  a  tax  on  real  and  personal  property,  §§ 58.1-3805 and 58.1-1718 authorize cities and counties to  impose  a  tax  on  the  probate  of  every  will  or  grant  of  administration  equal  to  one-third  of  the  state  tax  on  such  probate or grant of administration. Currently, the state tax on wills and grants of administration is 10 cents per \$100 or a fraction of \$100 for estates valued at greater than $15,000 (§  58.1-1712).  Therefore,  the  maximum  local  rate  is 3.3 cents.
|
|   A  related  *state*  tax  is  levied  in  localities  associated  with  the  Northern  Virginia  Transportation  Authority.  The  tax is a grantor’s fee of \$0.15 per $100 on the value of real property property sold. This was created as part of the 2013 transportation bill.
|
|   **Table  18.1**  provides  information  on  the  recordation  tax  and  the  wills  and  administration  tax  for  the  35  cities  and 89 counties that report imposing one or both of them. The following text table shows range of recordation taxes and taxes on wills and administration imposed by localities. 

```{r}

#Text table "Recordation Tax and Tax on Wills and Administration, 2019" goes here

#Table 18.1 "Legal Document Taxes, 2019" goes here
```


<!--chapter:end:va-local-tax-rates-Section-18.Rmd-->


# Section 19

Placeholder


## Miscellaneous Taxes, 2018
### LOCAL OPTION AND STATE SALES AND USE TAXES
### STATE MOTOR FUELS TAX ON DISTRIBUTORS
### BANK FRANCHISE TAX
### COMMUNICATIONS SALES AND USE TAX
### SHORT-TERM DAILY RENTAL TAX

<!--chapter:end:va-local-tax-rates-Section-19.Rmd-->




<!--chapter:end:va-local-tax-rates-Section-2.Rmd-->


# Section 20

Placeholder


## Refuse and Recycling Collection Fees, 2019
### REFUSE COLLECTION
### RECYCLING PROGRAMS

<!--chapter:end:va-local-tax-rates-Section-20.Rmd-->


# Section 21

Placeholder


## Residential Water and Sewer Connection and Usage Fees, 2019
### CONNECTION FEES
### USAGE FEES

<!--chapter:end:va-local-tax-rates-Section-21.Rmd-->


# Section 22

Placeholder


## Impact Fees for Roads, 2019

<!--chapter:end:va-local-tax-rates-Section-22.Rmd-->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
```

# Section 23

## Public Rights-of-Way Use Fees, 2019
The Code of Virginia § 56-468.1 authorizes certain localities to charge rights-of-way use fees for the use of publicly owned roads  and  property  by  certified  telecommunication firms. Cities and towns whose public streets are not maintained by the Virginia Department of Transportation (VDOT), as well as any county that has chosen to withdraw from the secondary system of state highways (currently only Arlington and Henrico counties), may impose a public rights-of-way use fee by local ordinance. This fee is in exchange for the use of the locality’s lands for electric poles or electric conduits by certified providers of telecommunications services. 

The provider collects the use fee on a per access line basis by adding the fee to each end-user’s monthly bill for local exchange telephone service (§ 56-468-1.G). The fee must be stated separately on the phone bill.

The fee is calculated each year by VDOT based on information  about  the  number  of  access  lines  and  footage  of  new  installation that have occurred in the reporting localities. Based on  this  information,  VDOT  uses  a  formula  to  calculate  the  monthly fee per access line for participating localities. Starting July  1,  2019,  the  fee  was  $1.20  per  access  line.  Information  about the rights-of-way use fee can be obtained from VDOT at: http://www.virginiadot.org/business/row-usefee.asp. The Code(§ 56-468.1.I) also permits any locality which had a franchise agreement or ordinance prior to July 1, 1998 to “grandfather” in the prior agreement provided that the county, city, or town does not discriminate among telecommunications providers and does not adopt any additional rights-of-way practices that do not comply with current laws.

***Table 23.1*** lists the localities that report having a rights-of-way agreement or a prior agreement that has been grand-fathered. The information is based on the Cooper Center’s 2019 survey. The text table below summarizes the results:   

```{r}
#Text table "Public Rights-of-Way Use Fees, 2019" goes here

#Table 23.1 "Localities Imposing Public Rights-of-Way Use Fees, 2019*" goes here
```

-----------------------------------------------------------------------------------------------------------

* In years prior to 2009 this table was based on information provided by the Virginia Department of Transportation. The current table uses data based on responses to the Cooper Center’s survey. To compare survey responses with VDOT information, refer to http://virginiadot.org/business/row-usefee.asp

<!--chapter:end:va-local-tax-rates-Section-23.Rmd-->


# Section 24

Placeholder


## Cash Proffers, FY 2018

<!--chapter:end:va-local-tax-rates-Section-24.Rmd-->


# Section 25

Placeholder


## Virginia Enterprise Zone Program, 2018*
### INTRODUCTION
### PURPOSE FOR THE PROGRAM
### PROGRAM GRANTS 
### LOCAL INCENTIVES
##### Source: Virginia Department of Housing and Community Development, Grant  Year  2018  Annual  Report:  Virginia  Enterprise  Zone  Program. Provided by the DHCD to the author.

<!--chapter:end:va-local-tax-rates-Section-25.Rmd-->


# Section 26

Placeholder


## Fiscal Content Information on Local Web Sites, 2019

<!--chapter:end:va-local-tax-rates-Section-26.Rmd-->



## Real Property Tax Relief Plans and Housing Grants for the Elderly and Disabled, 2019

<!--chapter:end:va-local-tax-rates-Section-3.Rmd-->

