# Fiscal Content Information on Local Web Sites

Because  the  web  is  such  an  inexpensive  way  to  provide  fiscal information, it has moved from being a backup source to a primary source. For that reason, we include a section in  the  survey  asking  localities to  provide  information  on  what  budget,  financial  and  tax  information  they  carry  on the web.

The first question was about the existence of a locality web  site.  If  the  answer  was  affirmative,  then  we  were  interested  in  knowing  if  the  locality  carried  information  about its budget, tax rates, capital programs, utilities, land book,  geographic  information  system  (GIS)  mapping,  and  audit  (technically  called  the  Comprehensive  Annual  Financial  Report  or  CAFR).  There  were  eight  questions  about these topics.

***Table 26.1*** lists the answers from the respondents. The text  table  summarizes  the  fiscal  content  information  for  those localities that answered affirmatively the question of whether there was a web site.

All cities and counties have web sites. Of the towns, 108 that answered the survey had a site. Many more localities maintain a web site now than in 2003, the first year we asked for information about web sites. In that year only 18 cities, 26 counties and 19 towns reported they had a web site. 

Currently, 32 cities and 79 counties, about  three-fourths of each, show web information on their proposed budget. Forty-nine towns reported having the proposed budget on their  sites.  Higher  numbers  of  cities,  counties,  and  towns  reported showing adopted budgets on the web, with 37 cities, 89 counties, and 80 towns reporting listing them.

Large majorities of cities (36), counties (91), and towns (94) with web sites showed tax rates. Utility rate schedules were shown by 34 cities, 49 counties and 87 towns. Not all jurisdictions maintain their own systems, a fact that should be considered in evaluating web sites. Capital improvement programs are shown by 35 cities, 54 counties, and 28 towns. In many cases capital programs may be reported as part of the adopted budget instead of as a separate category.

```{r}
library(tidyverse)
library(knitr)
```


```{r echo=FALSE,results='asis'}

#table name: Resources Available on Locality Websites, 2019
#SQL database connection for appropriate info replaces the 0's in these lists
item <- 0 
cities <- 0
counties <- 0
towns <- 0
total <- cities+counties+towns

table <- tibble('Item' = item, "Cities" = cities,"Counties" = counties,"Towns" = towns,"Total"=total)
kable(table,caption="Resources Available on Locality Websites, 2019")
```

By law, all localities must provide public access to the land book — the local listing of individual land parcels by owner and the assessed value of the land and improvements. Such  access  is  greatly  enhanced  when  it  can  be  provided  on the web. A majority of cities (34) and counties (84) now provide convenient web access to this important information. Most  with  web  access  also  provide  corollary  geographic  information system (GIS) mapping. Relatively few towns reported web inclusion of the land book or GIS mapping, a reflection of the fact that towns generally rely on their host counties for real property assessments. 

Almost  three-fourths  of  the  cities  and  over  half  the    counties  with  web  sites  reported  showing  their  latest  comprehensive annual fi nancial report (CAFR). Thirty-two cities, 61  counties, and 44 towns reported doing so.

```{r echo=FALSE, results='asis'}

#Table 26.1

#SQL query results will replace placeholder values (modified as necessary, depending on how the info comes out)

locality <- 0
website <- 0
budgetP <- 0
budgetA <- 0
rates <- 0
improvementProg <- 0
utilities <- 0
landbook <- 0
GIS <- 0
audit <- 0

table26.1 <- tibble("Locality" = locality,"Have A Website?" = website, "Proposed Budget" = budgetP,"Adopted Budget" = budgetA, "Tax Rates" = rates, "Capital Improvement Programs"=improvementProg,"Utility Charges"=utilities,"Landbook Information" = landbook,"GIS Mapping" = GIS, "Audit (CAFR)" = audit)


kable(table26.1,caption="Table 26.1 Fiscal Content Information on Local Websites, 2019")

```

