# Introduction {-}

## Foreward  {-}
This  is  the  thirty-eighth  edition  of  the  Cooper  Center’s  annual publication about the tax rates of Virginia’s local governments. In addition to information about tax rates, the publication contains details about tax administration, valuation methods, and due dates. There is also information on water and sewer rates, waste disposal charges and numerous other aspects  of  local  government  finance. This comprehensive guide to local taxes is based on information gathered in the spring, summer, and early fall of  2019. The study includes all of Virginia’s 38 independent cities and 95 counties and 118 of the 190 incorporated towns. The included towns account for 92 percent of the Commonwealth’s population in towns$^1$. The  study  also  contains  information  from  several  outside  sources, including two Department of Taxation studies, 2019 Legislative Summary and The 2017 Assessment/Sales Ratio Study, as well as Department of Taxation information on the assessed value of real estate by type of property. We also used the Auditor of Public Accounts’ Comparative Report of Local Government Revenues and Expenditures, Year Ended June 30, 2018, the Commission on Local Governments’ Report on Proffered Cash Payments and Expenditures by Virginia’s Counties,  Cities  and  Towns,  2017-2018,  and  the  Depart-ment of Housing and Community Development’s Virginia Enterprise Zone Program 2018 Grant Year Annual Report.


## Organization of the Book {-}

The study is divided into 26 sections. Section 1 is a reprint of the “Local Tax Legislation” section of the Department of Taxation’s 2019 Legislative Summary. The original Department of Taxation report is available at its website$^2$. Sections 2 through 26 cover specific taxes, fees, service charges, cash proffers, enterprise zones, and financial documents on the web. Most of the data came from a detailed web-based ques-tionnaire sent to all cities, counties, and incorporated towns (see Appendix A for a printed version). Appendix B provides a  listing  of  names,  phone  numbers,  and  email  addresses,  when available, of respondents and non-respondents to the questionnaire. Appendix C shows the percentage share of total  local  taxes  represented  by  each  specific  tax  for  each  locality based on data from the Auditor of Public Accounts for fiscal year 2018. Information is provided for each city and county and for 38 populous incorporated towns. Finally, Appendix D contains 2018 population estimates for cities, counties and towns  from the Cooper Center’s Demographics Research Group. The population information is provided to  give  readers  some  perspective  on  the  relative  size  of  localities. 

Please  note  that  the  web  addresses  provided  in  this  publication  were  good  at  the  time  this  text  was  printed.  However, some links are unstable and may not work with certain  browsers  or  they  may  be  modified  or  withdrawn  subsequent to publication. 

## About the Survey {-}

In 2019, localities could choose between online or printed versions of the questionnaire. The Cooper Center has made its best efforts to accurately refl  ect in this report the responses of localities based on the survey or follow-up queries.

In the tables three dots (...) are used to show there was no response and “N/A” is used to indicate “not applicable.” Readers may use the telephone/email list in Appendix B to contact  local  offi  cials  in  order  to  obtain  clarification  and additional detail. 

## Some Components of Local Taxes {-}

This  book  deals  mainly  with  local  sources  of  revenue  for  local  governments.  Though  localities  might  also  receive  federal and state resources, an important part of local funding comes from local sources. The Auditor of Public Accounts, Comparative  Report  of  Local  Government  Revenues  and  Expenditures provides data on these local sources. The fol-lowing analysis uses the data from their report for the year ended June 30, 2018.

A  common  misperception  is  that  nearly  all  local  tax  revenue  comes  from  the  real  property  tax.  True,  the  real  property  tax  is  the  dominant  source,  accounting  for  61.9  percent of city-county tax revenue in FY 2018, the latest year available (see text table below). But three other taxes—the personal  property  tax,  the  local  option  sales  and  use  tax,  and the business license tax—together accounted for 24.5 percent of total tax revenue. The remaining 14.6 percent of tax revenue came from more than a dozen other taxes. 

```{r, include=FALSE}
library(tidyverse)
library(knitr)
```

```{r echo=FALSE, results='asis'}
#Insert "Sources of Virginia Local Government Tax Revenue, FY 2018" here
tax <- c("Total taxes", "Real property","Personal property","Local option sales and use","Business license","Restaurant meals","Public service corporation property","Consumer utility","Hotel and motel room","Machinery and tools","Motor vehicle license","Recordation and will","Bank stock","Other local taxes","Tobacco","Coal, oil, and gas","Admission","Franchise license","Merchants' Capital","Penalties and interest")

amount <- c('$17,967,385,766', '$10,946,877,675', '$2,370,758,768','$1,239,855,163','$771,958,263','$612,940,580','$412,121,081','$327,627,947','$244,412,96','$233,076,157','$197,705,384','$126,458,487','$117,199,137','$92,124,397','$65,150,996','$28,510,002','$21,815,169','$16,362,103','$14,301,188','$128,130,305')

percent <- c(100.00, 60.93,13.19,6.90,4.30,3.41,2.29,1.82,1.36,1.30,1.10,0.70,0.65,0.51,0.36,0.16,0.12,0.09,0.08,0.71)
table1<-tibble("Tax" = tax, "Amount ($)" = amount, "% of Total"=percent)

kable(table1,caption="Sources of Virginia Local Government Tax Revenue, FY 2018")
```

There are six localities where the real property tax is not  dominant.  Bath  and  Surry  counties  have  large  power  plants  that  pay  public  service  corporation  property  taxes  that  overwhelm  other  sources.  Buchanan  County  has  rich  mineral deposits subject to local severance taxes that exceed the real property tax. Covington City and Alleghany County receive large shares of revenue from machinery and tools taxes on  MeadWestvaco’s large paperboard manufacturing facility. Finally, the small city of Norton, the least populous independent city in Virginia$^3$  (3,908 in 2018) received almost as much money from the local option sales and use tax as from the real property tax. In the remaining 127 cities and counties where the real property tax is dominant, its relative importance  varies  from  30.3  percent  of  total  tax  revenue  in  Galax  City  to  78.8  percent  in  Lancaster  County  (see  Appendix C). 

Thirty-six cities (two cities–Hopewell and Petersburg–did  not  provide  information  for  the  2018  Comparative Report)  and 95 counties imposed four of the taxes shown in  the  previous  table—the  real  property  tax,  the  personal  property tax, the local option sales and use tax, and the public service corporation property tax. Most, but not all, localities imposed recordation and will taxes, consumer utility taxes, motor  vehicle  license  taxes,  and  taxes  on  manufacturers’  machinery and tools. Nonetheless, as shown in the next text table, there are a number of taxes, a few of them signifi cant sources  of  revenue,  which  are  not  levied  by  all  localities.  Also,  some  of  the  taxes  are  used  so  sparingly  that  their  revenue yield is very low.

```{r echo=FALSE, results='asis'}
#Insert "Number of Virginia Localities Imposing Taxes by Type, FY 2018" table here
tax2 <- c("Real property","Personal property","Local option sales and use","Public service corporation property","Consumer utility","Recordation and wills","Motor vehicle license","Machinery and tools property","Bank stock","Hotel and motel room","Business license","Restaurant meals","Franchise license","Merchants’ capital","Tobacco","Admission","Coal, oil, and gas","Other local taxes")
cities <- c(36,36,36,36,36,32,32,31,36,32,36,36,11,1,29,18,1,23)
counties <- c(95,95,95,95,92,93,86,85,64,67,52,49,37,43,2,3,6,49)
total <- cities + counties
table2 <- tibble("Tax" = tax2, "Cities" = cities, "Counties" = counties, "Total" = total)

kable(table2,caption="Number of Virginia Localities Imposing Taxes by Type, FY 2018")
```

There are three major reasons why local governments do  not  to  impose  some  taxes:  (1)  The  locality  lacks  a  tax  base for a particular tax (e.g., a locality must have a bank in order to apply a bank stock tax and a locality must have taxable mineral deposits to impose coal, oil, and gas taxes). (2) The locality is faced with state restrictions (e.g., county excise taxes on hotel and motel room rental have tax rate restrictions  imposed  by  the  state;  county  restaurant  meals  taxes must be approved in a voter referendum; tobacco taxes are permitted in only two counties; and county admissions taxes are subject to many restrictions). In regard to the busi-ness,  professional, and occupational  license tax (BPOL tax), counties must choose either the BPOL tax or the merchants’ capital tax. Counties are not permitted to impose a business license tax within the boundaries of an incorporated town situated within the county without permission of the town. This  means  that  counties  with  large  shares  of  business  activity within towns are motivated to impose a merchants’ capital tax that can be applied countywide. (3) The locality chooses not to impose a permitted tax (e.g., Richmond City, a community with a large cigarette manufacturing plant, has not adopted a consumer tobacco tax even though all cities are granted the authority to levy such a tax). 

## Partnership with LexisNexis {-}

The Weldon Cooper Center for Public Service is partner-ing  with  the  publisher  LexisNexis  to  produce  the  annual  Tax  Rates  books.  The  Cooper  Center  still  prepares  and  distributes the survey and writes up the results. LexisNexis  publishes the book and fulfi  lls orders from interested parties. This arrangement allows us to concentrate on providing the most accurate and up-to-date information about Virginia tax rates  and  leverages  LexisNexis’  considerable  expertise  in  production and distribution of the annual volume. We hope the arrangement will lead to continued improvements in our Virginia Local Tax Rates series. 

## Study Personnel {-}

Stephen  C.  Kulp,  Research  Specialist  at  the  Center  for  Economic and Policy Studies, was responsible for work on the project. He refined the new database, administered the survey, translated the results into tables, checked relevant code sections, assisted with the development of the web-based questionnaire, and made appropriate changes in the text. Jennifer Nelson, of the Cooper Center’s Publications Section, designed the cover. Cooper Center employee Albert W. Spengler, who authored this study for a number of years prior to 1991, laid the foundation for the study when it was his responsibility.
 
The strong support for this publication by the Virginia Association of Counties and the Virginia Municipal League helps ensure our continued efforts to provide this resource as a basic reference on Virginia local taxes.

## Final Comments {-}

The Cooper Center is grateful to the many local officials throughout  the  commonwealth  who  supplied  the  survey information  presented  in  this  study.  Their  willingness  to  provide information and their patience in answering follow-up questions is what makes this book successful. The high response rates could not have been achieved without their cooperation. Corrections to the text or suggestions for possible changes or additions to future editions can be made using the email address and phone number listed below.

Stephen C. Kulp

Research Specialist

Center for Economic and Policy Studies

Weldon Cooper Center for Public Service

University of Virginia

Charlottesville

February 2020



$^1$  Locality  population  figures are based on estimates developed by the Demographics Research Group of the Weldon Cooper Center for Public Service. See Appendix D.

$^2$  https://tax.virginia.gov/legislative-summary-reports

$^3$  Weldon  Cooper  Center  for  Public  Service,  University  of  Virginia.  https://demographics.coopercenter.org/population-estimates-age-sex-race-hispanic-towns/
