---
title: "va-local-tax-rates Section 2"
author: "Anne Bader"
date: "6/17/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Section 2
#Real Property Tax, 2019

The real property tax is by far the most important source of tax revenure for localities. In fical year 2018, the most recent year available from the Auditor of Public Accounts, it accounted for 55.5 percent of tax revenue for cities, 64.6 percent for counties, and 29.1 percent for large towns. These are averages; the relative importance of taxes in individual cities, counties, and towns varies signifigantly. For information on individual localities, see Appendix C. 
The Code of Virgina, §§ 58.1-3200 through 58.1-3389, authorizes localities to levy taxes on real property (land, including the buildings and improvements on it). There is no restriction on the tax rate that may be imposed. Section 58.1-3201 provides that all general reassessments or annual assessments shall be at 100 percent of fair market value. 

#PUBLIC SERVICE CORPORATIONS

Property owned by so-called public service corporations is
not assessed by localities. Instead, that task is delegated to
the State Corporation Commission (SCC) and the Department of Taxation.The State Corporation Commission
assesses electric utilities and cooperatives, gas pipeline
distribution companies, public service water companies,
and telephone and telegraph companies. The Department
of Taxation assesses pipeline transmission companies and
railroads.

In fiscal year 2018, the property tax on public service
corporations accounted for 1.7 percent of tax revenue for
cities, 2.6 percent for counties, and 0.8 percent for large
towns. These are averages; the relative importance of the tax
in individual cities, counties, and towns varies significantly.
In two counties with large power generating facilities the
property tax on public service corporations accounts for a
very large share of local tax revenue. In Bath County the
share was 47.6 percent and in Surry County the share was
61.1 percent. For more information on individual localities,
see Appendix C.

The commissioner of the revenue or another designated
official in each city or county is required to provide by
January 1 of each year to any public service company with
property in its area a copy of the property boundaries of
the locality in which any part of the company is located
(§ 58.1-2601). The State Corporation Commission or the
Department of Taxation send out their assessments for the
property based on these boundaries (§ 58.1-2602). Localities examine the assessments to determine their correctness.
If correct, the locality determines the equalized assessed
valuation of the corporate property by applying the local
assessment ratio prevailing in the locality for other real estate (§ 58.1-2604). Local taxes are then assigned to real and tangible personal property at the real property tax rate current in the locality (§ 58.1-2606).

#TAX RELIEF PROGRAMS

There are several types of locally financed tax relief programs available. Section 3 of this study contains information
on so-called circuit breaker plans for the elderly and disabled. Section 4 covers land use assessments for agricultural,
horticultural, forestal, and open space real estate. Section
5 contains information on preferential assessments for
agricultural and forestal districts. Finally, Section 6 covers
property tax exemptions for certain rehabilitated real estate
and other exemptions.

Only the city of Charlottesville, Loudoun County, and
Arlington County reported providing tax relief for low-income
owners and renters who are not elderly. The city of Charlottesville administers the Charlottesville Housing Affordability Program (CHAP) to help low and middle income homeowners. The program awards grants up to $1,000 to homeowners
with houses assessed at less than $375,000 and having an
annual income less than $55,000.$^1$ Loudoun County administers the Affordable Dwelling Unit Program for renters and
first-time buyers. Buyers need an income greater than 30
percent but less than 70 percent of the area median income to
participate. Qualified renters are eligible to rent apartments at rates from $630 to $1,300. Arlington County’s Housing Grants
Program is available to working families with at least one child
under age 18. Personal assets may not exceed $35,000 and
there is an income limit based on household size.

Localities are permitted to institute deferral for a portion of the real estate tax by § 58.1-3219 of the Code of
Virginia. Localities are permitted to grant deferrals from
the full amount by which each taxpayer’s real estate tax
levy exceeds 105 percent of the previous year’s tax, or
such higher percentage adopted by the locality. Deferred
taxes are subject to interest in an amount established by the
governing body, not to exceed the rate published by the
IRS code.$^2$ The deferral potentially applies to every property
owner, not just the elderly and disabled. (For deferrals limited
to the elderly and disabled see Section 3 of this study.)

The deferral program is rarely used. Administrative problems appear to be the major reason for the unpopularity of deferral programs. Loudoun County had a deferral program in place
in the 1990s but terminated it “… because the program was
administratively complex, cumbersome and required
staff time in disproportion to the benefit received by the
taxpayer.”$^3$ The cities of Alexandria, Falls Church, and Fairfax
and the counties of Fairfax and Henrico considered deferral but
did not adopt it. According to Henrico staff, “The administrative procedures for tracing the properties and recovering the
relevant taxes upon either the death of the owner or transfer
of the property itself would be both cumbersome and time
consuming and could not be accomplished with existing
staffing levels or existing computer systems.”$^4$ Another
reason for the unpopularity of the programs may be that
taxpayers only receive postponement, not removal, of the
tax liability. The cities of Charlottesville and Richmond, the
county of Middlesex, and the town of Amherst were the only
localities reporting a deferral program in 2019. 

#STATUTORY RATES, SPECIAL TAXES, DUE DATES, PRORATION, AND BILLING PRACTICES

Table 2.1 provides general information associated with real
property taxes in Virginia’s localities. The table provides
an estimate by locality of both the number of total taxable
real estate parcels and the number of residential parcels.
Twenty-seven cities, 80 counties and 52 towns provided
estimates of one or both types of parcels. The total number
of parcels in cities ranged from a high of 158,431 (Virginia
Beach) down to 2,456 (Lexington). Among counties, the
number of parcels ranged from a high of 354,687 (Fairfax)
down to 3,940 (Highland).

Table 2.1 also lists the statutory (nominal) tax rates. The
statutory rate is the rate used by localities and is applied
to the assessed value of a property. In the table, statutory
rates are listed under calendar year (CY) or fiscal year (FY)
columns depending on the locality’s assessment cycle. In
most cases the calendar year tax rate listed runs from January 1 to December 31 and the fiscal year rate runs from
July 1 to June 30. The provisions explaining the assessment
cycle requirements are found in § 58.1-3010 and § 58.1-3011
of the Code of Virginia. However, some localities report a
calendar year assessment schedule with a fiscal year valuation. Six cities (Chesapeake, Harrisonburg, Martinsville,
Roanoke, Salem, and Suffolk) and one county (James City)
report this practice. Otherwise, 15 cities and 88 counties
reported using the calendar year cycle while 176 cities and
6 counties used fiscal year assessment cycles. 

The statutory tax rates were reported to the
Cooper Center by all cities and counties and 112 of the
responding towns. The text table below lists the averages
for the statutory rates from the localities. 

```{r}

#table name: Statutory Real Estate Tax Rates per $100 of Assessed Taxable Value for Localities Reporting, CY 2019 and FY 2020


```

Statutory rates are generally higher for cities than
counties. The rates are lowest in towns because they are
subordinate to counties and have limited responsibilities.

Tax due dates vary among localities. Generally,
if taxes are paid annually, they are due by December
5. If paid semiannually, they are due by June 5 and
December 5. However, some localities have different due
dates, as provided by § 58.1-3916 of the Code. 

Most cities have semiannual tax due dates with payments required in June and December. Of the 38 cities,
2 required taxes due annually, 31 semiannually, and 5
quarterly. Among the counties, 32 had annual tax due
dates, while 63 had semiannual requirements. Of the towns
responding to this question, 80 reported annual due dates,
and 32 required semiannual payments.

A locality is permitted to prorate the taxable amount.
Any county, city, or town electing to prorate new buildings
which are substantially complete prior to November 1 must
do so at the time the building is complete or fit to live in. Of
the 38 cities, 33 reported prorating taxes while 5 reported
not doing so. Among counties, 67 prorated their taxes while
28 did not. Reports from the towns that answered this question indicated that 47 prorated their taxes while 652 did not.

The final column of Table 2.1 pertains to town billing
practices. Three possibilities exist: (1) a town sends out its
own bills and collects its taxes (TT in the table), (2) a town
collects its taxes but the county sends the bills (CT in the
table), or (3) a town has the county bill and collect the taxes
(CC in the table). Of the towns that answered the question,
the overwhelming majority, 100, reported billing and collecting their own taxes. Four said they collected taxes, while
in three the county both billed and collected town taxes.

Table 2.2, Table 2.3, and Table 2.4 provide additional
information concerning statutory real property tax rates.
The Code allows localities to add special purpose levies on
top of the real property rate for various purposes. Table 2.2
deals with the category of special districts. A special district
is organized to perform a single governmental function or a restricted number of related functions. Special districts
usually have the power to incur debt and levy taxes to fund
special activities such as capital improvements, emergency
services, sewer and water services, or pest control within
those districts. Thirteen cities, 14 counties, and 4 towns
reported levying these taxes. The table includes the base
(statutory) rate for the locality, the district in which the activity takes place, the purpose of the activity, and the special
rate imposed for that activity. Most special activity taxes are
in addition to the base rate, though some are simply a flat
fee, and others are a percentage rate based on improvements
to the property.

Another special district category is the community
development authority (CDA). Such an authority is a district created by the locality based on a petition from the
property owners to help develop and maintain desired public
infrastructure improvements, such as roads and buildings.
The CDA is usually associated with development interests,
such as retail centers, industrial centers, or tourism centers.
Generally the CDA pays for development by issuing bonds
and then having the property owners pay special assessments
based on the level of debt. Assessments are levied either by
placing a tax, such as $0.25 per $100 of assessed value, on
the property within the district or by a special assessment
each year that determines the benefit from the improvements
and allocates them by property value. Depending on how the
bond agreement is structured, assessment payments may be
made directly to bondholders or to the locality. Table 2.3 lists
community development authorities by locality. The table
includes the name of the project, the purpose, the size, the
bond amount, and, where possible, the current value. Three
cities and 8 counties reported having CDAs. 

The final category of special districts is that of localities within the Northern Virginia Transportation Authority.
Localities within this authority have the ability to tax real
property associated with industrial and commercial use up
to $0.125 per $100 of assessed value to help fund transportation improvements. In 2009, an amendment to § 58.1-3221.3
specified that the revenues generated by the tax were to be
used solely for (1) new road construction, design, and right-of-way acquisition, (2) new public transit construction,
design, and right-of-way acquisition, (3) capital costs related
to new transportation projects, or (4) the issuance costs and
debt service on any bonds issued to support capital costs.
There are 11 localities in the region of the authority: the
cities of Alexandria, Fairfax, Falls Church, Fredericksburg,
Manassas, and Manassas Park and the counties of Arlington,
Fairfax, Loudoun, Prince William, and Stafford. Of those,
one city (Fairfax) and two counties (Arlington and Fairfax)
reported implementing the tax, as shown in Table 2.4.

#ASSESSMENT PRACTICES, REASSESSMENTS, ASSESSED VALUES

Table 2.5 details assessment practices among
localities. The table includes cities and counties, but not
towns, because only a small percentage of towns provided
substantive answers. For those interested in the towns that responded, data are available from the Cooper Center
upon request. 

The second column lists whether a locality has a full-time assessor. Twenty-seven cities reported employing a full-time property tax assessor, while 11 did not. In contrast, only
36 counties had a full-time assessor while 59 did not. This
reflects the fact that many counties reassess property less
frequently than cities. No towns had assessors, since towns
rely on assessed values established by their host counties. 

Columns three, four, five, and six of Table 2.5 provide
data on the conduct of general reassessments and cover
four questions. (1) Are reassessments done by the locality
or contracted out? (2) What is the reassessment frequency?
(3) Is physical inspection part of the reassessment? (4)
When was the reassessment last done? Regarding the
conduct of the general reassessment, 28 cities reported
conducting reassessments in-house while 10 reported
contracting with outside assessors. Twenty-eight counties
reported doing general reassessments in-house, while 67
reported contracting out for services. Section 58.1-3250
of the Code requires cities to have a general reassessment
of real estate every two years. However, any city with a
total population of 30,000 or less may elect to conduct its
general reassessments at four-year intervals.$^5$ Counties are
required to have a general reassessment every four years
(§ 58.1-3252). There is an exception for counties with a
total population of 50,000 or less. These counties may
elect to reassess at either five-year or six-year intervals
(§ 58.1-3252). However, nothing in these sections affects the
power of cities and counties to reassess more frequently. A
large majority of the cities (30) reassess at one or two year
intervals. In contrast, less than three out of ten counties (27)
reassess that frequently. Virtually all of the populous cities
and counties reassess annually or biennially. Towns rely
on their surrounding county to provide assessments, so a
town’s reassessment occurs with the same frequency as the
county’s. The reassessment periods are summarized in the
table on the following page.

Column seven of Table 2.5 shows information
about maintenance assessments. While general reassessments involve reassessing all parcels to reflect
changes in market value, maintenance assessments
involve adjusting assessed values between reassessments
because of new construction, improvements, damages,
demolitions, subdivisions, and consolidations. Thirty-three cities responded that they performed maintenance
assessments using staff, while five reported contracting
for the work. Among counties, 66 reported performing
maintenance reassessments using staff, while 29 reported
contracting the work to independent appraisers.

Columns eight and nine of Table 2.5 cover physical inspection. Physical inspection refers to the actual
inspection of the property as opposed to computerized
mass-appraisal of parcels. If a locality responded that
it did not perform physical inspections during the general reassessment, two further questions were asked: 

```{r}

#table name: Reassessment Periods for Real Estate, 2019


```


(1) Does the locality perform a physical inspection at all?
(2) If so, what is the inspection cycle? Among cities that
responded, 18 reportedly did not have a physical inspection
separate from the general reassessment cycle. Twenty others reported having a physical inspection cycle, the periods
ranging anywhere from two to six years. Among counties
that responded, 70 indicated they performed physical
inspections during general reassessment, while 25 reported
having physical inspection cycles ranging anywhere from
one to six years.

Table 2.6 provides unpublished Department of Taxation
2018 data on total taxable assessed value of real estate by
category. Taxable assessed value shows property qualifying for use value at its use value, not its market value. The
percentage distribution of taxable assessed value is shown
for two types of residential property (single-family and
multi-family) as well as commercial and industrial property
and agricultural property. 

The text table on the next page compares the taxable
assessed value by category for cities and counties. The total
assessed value for all cities amounted to $277.4 billion.
Single-family residential property averaged 64.9 percent
of taxable assessed value. Multi-family residential property
averaged 11.1 percent of taxable assessed value. Commercial/industrial properties averaged just over one-quarter of
the total value at 23.9 percent, while agricultural property
values amounted to only 0.1 percent.

The total assessed value of property by category for
counties in 2018 amounted to $854.5 billion. Of that amount,
72.0 percent of assessed value was associated with single-family residential property, 5.9 percent with multi-family
residential property, 18.2 percent with commercial/industrial
property, and 4.0 percent with agricultural property.

With the total amounts from cities and counties combined, the total assessed valuation amounted to $1,131.9 billion. Of that, 70.2 percent applied to single-family residential
property, 7.1 percent applied to
muli-family residential property,
19.6 percent applied to commercial/industrial property, and 3.0
percent to agricultural property.

Looking at the percentage
breakdown for each type of locality, in 2018 the share of taxable

```{r}

#table name: Taxable Assessed Value by Category for Cities and Counties, 2018

```

assessed value for cities in the single-family residential category was between 40 percent and 59.9 percent in 19 cities
and 60 percent or more in 18 cities. All cities but two had
multi-family residential values under 19.9 percent of the
total assessed value. Commercial and industrial property
was the second most common category with 21 of the cities
having between 20 percent and 39.9 percent of their property
valuations coming from this type of property. Finally, only
the cities of Suffolk and Franklin had more than 2 percent
of their property valuation associated with agriculture.

Among counties the breakdown was slightly different.
As in cities, the single-family residential value dominated the
percentage breakdown. The single-family residential assessment percentage amounted to 60 percent or more for 71
counties. Another 20 received between 40 percent and 59.9
percent of the valuation from single-family residential real
estate, while in four counties residential valuations amounted
to no more than 39.9 percent of the total taxable assessed
value (Buchanan, Dickenson, Highland, and Sussex). In
contrast, only in Arlington county did the multi-family
residential average share of value exceed 19.9 percent. 

The category with the second highest valuation in
counties was commercial and industrial property. Eighty-two counties had such property valued no higher than 19.9
percent of the total assessed value of property within the
locality. In general, the percentage of assessed value in
counties for commercial and industrial properties was less
than that for cities (though two counties, coal-rich Dickenson
and Buchanan, had the highest percentage valuations of such
property). Finally, agricultural property averaged the least
total assessed valuation in counties, though the percentage
varied greatly among the individual counties. In 30 counties,
valuations associated with agricultural property made up 20
percent or more of the total assessed value within the locality.
The percentage in one county (Sussex) was 82.0 percent.
The taxable assessed values for agriculture were much lower
than they would have been without the advantage of use
value assessment, a program explained in Sections 4 and 5. 

#EFFECTIVE TAX RATES

Tax rates are generally discussed in terms of either statutory
(nominal) rates or effective rates. The statutory rate is the
rate used by localities and is applied to the assessed value
of a property.The effective rate is published by the Virginia
Department of Taxation in their annual assessment/sales
ratio study. The department derives the effective rate by
multiplying the statutory tax rate by the median assessment
ratio. In normal times when property values are rising, the
median assessment ratio is usually less than 100 percent 

```{r}

#table name: Share of Assessed Value of Real Estate by Category, 2018

```

because reassessments lag market increases and tend to be
conservative. Consequently, the statutory rate is generally
higher than the effective rate. However, this may not be true
in difficult real estate markets. A limitation of the effective
rates published by the Virginia Department of Taxation is
that they are not current. The most recent year available at
the present time is 2017. Despite the time lag, effective rates
are important because they give a more accurate reflection
of the differences in real property tax rates across localities.

Table 2.7 shows city and county average effective tax
rates in the year 2017. The department makes its computation in order to control for the variance in localities’ assessment procedures and timing. Therefore, when comparing
tax rates among localities, the reader may wish to consult
both Tables 2.1 and 2.7. Table 2.1 shows statutory rates in
2019. Table 2.7 shows statutory and effective rates in 2017.
The following text table summarizes the effective tax rates
for the localities shown in Table 2.7.

It should also be pointed out that the Virginia Department of Taxation does not use the locally reported statutory
tax rate in its computations. Instead, it calculates the statutory rate by dividing the real estate levy by the local real 

```{r}

#table name: Effective Real Estate Tax Rates, 2017

```

estate taxable assessed value,$^6$
as reported in the local land
book. This method of computing the statutory tax rate takes
additional district levies into account.$^7$

In 2 cities and 10 counties the statutory rate was less than
the effective rate. In two cities and seven counties statutory
and effective rates were the same. Finally, in 34 cities and
78 counties statutory rates exceeded effective rates. 

```{r}

#table name: Statutory and Effective Real Estate Tax Rates, 2017
```

The real property tax rates reported in Table 2.7 are a
more accurate reflection of the differences among localities
in tax rates on real property than those in Table 2.1 because
they control for variations in assessment frequency and
technique among localities. Table 2.7 also shows the latest
reassessment in effect when the median ratio study was
conducted, the number of sales used in the study, the median
ratio, and the coefficient of dispersion. 

The coefficient of dispersion measures how closely
the individual ratios of each locality are arrayed around the
median ratio. The formula for the coefficient of dispersion
(CD) is:

#Equation goes here

where Xi represents the assessment/sales ratio for the ith
sale in a sample of size n, and Xm represents the median
ratio of the sample.$^8$ If there were no dispersion, the CD
would equal zero.

The text table below summarizes the coefficients of dispersion tabulated for the cities and counties. Eighteen of the
cities had CDs of no more than 9.9 percent. Eight had CDs
between 10 percent and 14.9 percent, 7 had CDs between 15

```{r}

#table name: Coefficient of Dispersion, 2017

```

and 19.9 percent, and 4 had CDs between 20 and 24.9 percent.
Counties tended to vary more in the degree of dispersion.
Thirteen had CDs between 5 and 9.9 percent, 18 had CDs
between 10 and 14.9 percent, 25 had CDs between 15 and
19.9 percent, 26 had CDs between 20 and 24.9 percent, 11 had
CDs between 25 and 29.9 percent, and 2 had CDs between
30 and 34.9 percent. 

There is no upper limit for what is tolerable, but the
International Association of Assessing Officers recommends
an upper limit of 15 percent for residential properties.$^9$ Twenty-eight cities and 34 counties met the 15 percent standard.$^10$

As one would expect, the quality of local assessments,
as measured by the CD is generally better in those localities
that reassess annually, biennially, or that have just conducted
a general reassessment. In 2017, of the 57 localities with CDs
under 15 percent, all but 12 reassessed annually (28), biennially (10), or had just completed general reassessments (7).

#MISCELLANEOUS ITEMS

Table 2.8 presents miscellaneous taxes and exemptions
related to real property. The first is the recreation tax. The
Code in §15.2-1807 permits localities to collect a real
estate tax for recreation areas and playgrounds that is not
to exceed $0.02/$100 of the assessed value of a property.
This tax was reported by Charlottesville City. 

The second column refers to the tax deferral ordinance
permitted by § 58.1-3219 regarding the deferral of a portion
of real estate tax increases when the tax exceeds 105 percent
of the real property tax on property owned by a taxpayer
in the previous year. Four localities (Charlottesville City,
Richmond City, Middlesex County, and Amherst Town)
reported implementing this deferral. 

The third column refers to the establishment of a tax
increment financing fund used to encourage development in
certain areas and permitted by § 58.1-3245 of the Code. Six
cities (Bristol, Charlottesville, Chesapeake, Emporia, Newport News, Virginia Beach, and Waynesboro), four counties
(Arlington, Augusta, Fairfax, and Hanover), and one town
(Front Royal) reported having implemented such a fund. 

The fourth column refers to separate real property tax
rates for energy-efficient buildings as permitted by § 58.1-
3221.2 of the Code. Three cities (Charlottesville, Roanoke,
and Virginia Beach) reported having special rates for such
real estate. 

The fifth column lists localities that reported providing
a separate real property classification for improvements to
real property used in the manufacture of renewable energy.
Only the cities of Charlottesville and Roanoke reported
having this separate rate.

Finally, the last column refers to low-income grant
programs, discussed earlier in this text under the subheading, “Tax Relief Programs.” Only the cities of Charlottesville
and Norfolk, and the county of Arlington reported having these
programs. 

```{r}

#Table 2.1


```

```{r}

#Table 2.2

```

```{r}
#Table 2.3

```

```{r}

#Table 2.4

```

```{r}

#Table 2.5

```

```{r}

#Table 2.6

```

```{r}

#Table 2.7

```

```{r}

#Table 2.8

```



--------------------------------------------------------------------

$^1$ Charlottesville Housing Affordability Program: https://www.
charlottesville.org/departments-and-services/departments-a-g/
commissioner-of-revenue/real-estate-tax-relief-for-the-elderlyand-disabled. Loudoun County Affordable Dwelling Unit Program: http://www.loudoun.gov/adu. Arlington County Housing
Grants Program: http://housing.arlingtonva.us/get-help/rentalservices/local-housing-grants/.

$^2$  The statute allows the use of the Internal Revenue Service rate. Section 6621 of the Internal Revenue Code establishes a rate of 3 percent plus the federal short-term rate. In December 2019, when the short-term rate was 1.616 percent, the combined annual rate was 4.61 percent.

$^3$ City of Alexandria, Budget Memo #46: Review of Other Jurisdictions’ Experience with a Real Estate Tax Deferral Program for the General Population (Councilman Speck’s Request), 4/25/2003.

$^4$ Henrico County, Budget Memo #46.

$^5$ The Code does not specify which census is to be used.

$^6$ Taxable assessed value treats property qualifying for use value
as taxable at its use value rather than at its full market value.

$^7$ Virginia Department of Taxation, The 2017 Virginia Assessment/
Sales Ratio Study (Richmond, February 2019), p. 35. The study
can be found at https://tax.virginia.gov/assessment-sales-ratiostudies.

$^8$ Virginia Department of Taxation, The 2017 Virginia Assessment/
Sales Ratio Study, p. 34.

$^9$ International Association of Assessing Officers, Standard on
Ratio Studies, (approved April 2013), p. 17. http://www.iaao.
org/media/standards/Standard_on_Ratio_Studies.pdf.

$^10$ The Department of Taxation’s study applies to all types of property, not just residential property. Nonetheless, the majority of
the measured sales are for single-family residential properties. 

