
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Section 3
## Real Property Tax Relief Plans and Housing Grants for the Elderly and Disabled, 2019

Sections 58.1-3210 through 58.1-3218 of the *Code of Virginia* provides that localities may adopt an ordinance allowing property tax relief for elderly and disabled persons. The relief may be in the form of either deferral or exemption from taxes. The applicant for tax relief must be either disabled or not less than 65 years of age and must be the owner of the property for which relief is sought (§ 58.1-3210). The property must be the sole dwelling of the applicant. In addition, localities have the option of exempting or deferring the portion of a person’s tax that represents the increase in tax liability since the year the taxpayer reached 65 years of age or became disabled. 

Localities are allowed to establish by ordinance the net financial worth and annual income limitations pertaining to
owners, relatives and non-relatives living in the dwelling(§ 58.1-3212) of qualified elderly or handicapped persons.
Further, mobile homes that are owned by elderly and disabled persons are included in the allowable property tax
exemptions whether or not mobile homes are permanently affixed. Finally, local governments are authorized to extend
tax relief for the elderly and disabled to dwellings that are jointly owned by individuals, not all of whom are over 65
or totally disabled. 

The text table below indicates the range and media nof the combined gross income allowance and combined net worth limitations for those cities, counties, and towns responding to the survey. 

```{r}

#table name: Relief Plan Statistics: Gross Income and Net Worth, 2019
```

The following text table indicates, for those localities responding, how many localities have a tax relief plan that
applies to both the elderly and the disabled, the elderly only,or the disabled only. 

```{r}

# table name: Relief Plans for Elderly and Disabled, 2019
```

A majority of the localities exempt an owner from all or part of the taxes on the dwelling; usually the exemption is based on a sliding scale, with the percentage of the exemption decreasing as the income and/or net worth of the owner increases.

**Table 3.1** summarizes the various tax relief plans offered to elderly and disabled property owners in Virginia.
The figures under the combined gross income heading reflect, first, the maximum allowable income (including the
income of all relatives living with the owner) for an owner to be eligible for relief and, second, the amount of income
of each relative living in the household, except the spouse, who is exempted from this amount. 

For example, if the table reads “$7,500; first $1,500 exempt,” this indicates that the combined income of the
owner and all relatives living with him/her may not exceed $7,500, except that the first $1,500 of income of each relative other than the spouse is excluded when computing this amount. The combined net worth amount listed usually
excludes the value of the dwelling and a given parcel of land upon which the dwelling is situated. 

**Table 3.2** details relief plans for renters. As the table indicates, few localities offer such plans. Only five cities
(Alexandria, Charlottesville, Fairfax, Falls Church, and Hampton) and one county (Fairfax) reported having plans
for renters.

**Table 3.3** lists the combined elderly and disabled beneficiaries reported by each locality in 2018 or 2019 and
the amount of revenue foregone by each locality because of the homeowner exemptions. The amounts were reported
by 23 cities, 66 counties, and 31 towns that responded to the question. The amounts reported foregone totaled $21,698,890
for cities, $60,242,734 for counties and $636,229 for the reporting towns. The grand total amount foregone by
reporting cities, counties, and towns was $82,577,853. An estimate of the average revenue foregone per beneficiary
is also provided for localities reporting both number of beneficiaries and foregone revenue. For cities, the average
revenue foregone was $1,518 per beneficiary. The amount for counties was $1,581, and for towns it was $360. 

```{r}

#Table 3.1 Real Property Owner Tax Relief Plans for the Elderly and Disabled, 2019

#Table 3.2 Real Property Renter Tax Relief Plans for the Elderly and Disabled, 2019

#Table 3.3 Real Property Tax Relief Plans for the Elderly and Disabled Homeowners: Number of Beneficiaries and Foregone Tax Revenue, 2018 or 2019

```
